title: "R code for EdgeR transcriptome analysis"
author: "Shah Rukh"
date: "05/28/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

## Differential expression analysis with EdgeR - Cortex WT vs LgDel


```{r libraries}
rm(list=ls())
library(tidyverse)
library(edgeR)
library(factoextra)
library(plotly)
library(tidyr)
library(pheatmap)
library(statmod)
library(writexl)
library(biomaRt)
library(readxl)
library(DGEobj.utils)

```





Cleaning the data to generate counttablefull
```{r counttablefull}
files <- dir(path = "/Users/shahrukh/Library/CloudStorage/OneDrive-VirginiaTech/Downloads/cortical_culture_rnaseq/WT_NAC_Tab files", 
             pattern = "*.tab", full.names = T)

counttablefull <- files %>%
  map(read_tsv,  skip = 4, col_names = FALSE ) %>%
  purrr::reduce(cbind) 



datasets <-
  files %>%
  stringr::str_replace("/Users/shahrukh/Library/CloudStorage/OneDrive-VirginiaTech/Downloads/cortical_culture_rnaseq/WT_NAC_Tab files/", "") %>% # replace data/ at the beginning of filename
  stringr::str_replace("ReadsPerGene.out.tab", "") # replace ReadsPerGene.out.tab end of name
datasets

columnnames <- c()
for (i in datasets) {
  columnnames <- c(columnnames,
                   paste0("gene_", i),
                   paste0(i, "_unstranded"),
                   paste0(i, "_forwardstrand"),
                   paste0(i, "_reversestrand")
  )
}


names(counttablefull) <- columnnames
rm(columnnames, datasets)


counttablefull <- counttablefull %>%
  dplyr::mutate(ensembl_gene = gene_WT_2) %>%
  dplyr::select(-starts_with("gene"))

```

Test for strandedness and selecting strand specific column

```{r Test for strandedness}
counttablefull %>% # remove gene name column
  dplyr::select(-ensembl_gene) %>% # find column sum
  summarise_each(funs(sum)) %>%  # gather wide data (columns) into a long table
  gather(library, counts) %>% # split library column into dataset and protocol 
  separate(library, into = c("treatment", "replicate", "stranding"), sep="_") %>% # split library into 3 columns
  spread(stranding,counts) %>% # make columns for each of the strandings
  dplyr::mutate(propF = forwardstrand/unstranded, propR = reversestrand/unstranded) # assess rev/unst and for/unst


counttablefull %>% # remove gene name column
  dplyr::select(ends_with("unstranded")) %>%
  summarise_each(funs(sum)) %>%  # gather wide data (columns) into a long table
  gather(library, counts) %>%  # split library column into dataset and protocol
  separate(library, into = c("treatment", "replicate", "stranding"), sep="_") %>%
  dplyr::select(-stranding) %>%
  ggplot(aes(y = counts, x = treatment, fill = replicate)) +geom_bar(stat = "identity", position = "dodge")
#ggsave("count_by_gene_table.png")


counttablefull <- counttablefull %>% 
  dplyr::select(-ends_with("reversestrand"), -ends_with("forwardstrand")) 
names(counttablefull) <- str_replace(names(counttablefull), "_unstranded", "")
dim(counttablefull)
```




```{r plot the counttablefull}
counttablefull %>%
  dplyr::select(-ensembl_gene) %>%
  gather(library, value) %>%
  ggplot(aes(x = value)) + geom_histogram(bins = 100) + facet_wrap(~library)
#ggsave("genes_with_low_exp_hist.png")
counttablefull %>%
  dplyr::select(-ensembl_gene) %>%
  gather(library, value) %>%
  filter(value > 0) %>% 
  ggplot(aes(x = value)) + geom_density() + facet_wrap(~library) + scale_x_log10()
#ggsave("genes_with_low_exp_log.png")
counttablefull %>%
  dplyr::select(-ensembl_gene) %>%
  gather(library, value) %>%
  filter(value > 0) %>% 
  ggplot(aes(y = value)) + geom_boxplot() + facet_wrap(~library) 
#ggsave("genes_with_low_exp_box.png")
```


```{r Remove ensembl gene version character}
###remove gene version from ensembl ids####
counttablefull <- counttablefull %>% separate(ensembl_gene, into=c("ensembl_gene", "version"), sep = "\\.")
counttablefull <- subset (counttablefull, select = -version)
```



```{r Make counttable matrix and make cpm counttable}
counttablematrix <- counttablefull %>%
  dplyr::select(-ensembl_gene) %>%
  as.matrix()

row.names(counttablematrix) <- counttablefull$ensembl_gene
counttable_cpm <- cpm(counttablematrix)
head 
```



```{r filter counttable_cpm using cpm threshold cutoff and required sample number meeting cpmThreshold}

counttablefull %>%
  dplyr::select(-ensembl_gene) %>%
  purrr::modify(~sum(.)) %>% 
  distinct() %>%
  gather(library, counts) %>%  # split library column into dataset and protocol
  dplyr::mutate(millionreads = counts/(10^6)) %>%
  dplyr::mutate(cpmThreshold = 15/millionreads)  %>%
  dplyr::mutate(mycutoff = 0.35 * millionreads)

subsetting_matrix <- counttable_cpm > 0.35
head (subsetting_matrix)

subsetting_vector <- rowSums(subsetting_matrix) >= 3
head(subsetting_vector)

counttablematrix_filt <- counttablematrix[subsetting_vector,]
dim(counttablematrix_filt)
```



## Differential expression analysis starts
```{r create matrix and visualize unprocessed data}

mydgelist <- DGEList(counttablematrix_filt)
mydgelist

head(mydgelist$samples)

mydgelist$samples %>% 
  as.data.frame() %>%
  rownames_to_column() %>%
  separate(rowname, into = c("treatment", "replicate"), sep="_") %>%
  ggplot(aes(y = lib.size, x = treatment, fill = replicate)) +geom_bar(stat = "identity", position = "dodge")
#ggsave("count_by_gene_filtered.png")

mydgelist_cpm <- cpm(mydgelist, log = TRUE)
boxplot(mydgelist_cpm)
abline(h=median(mydgelist_cpm),col="blue")
title("Boxplots of logCPMs (unnormalised)")

mydgelist_cpm %>% 
  as.data.frame() %>%
  rownames_to_column() %>%
  gather(library, log2cpm, -rowname) %>%
  separate(library, into = c("treatment", "replicate"), sep = "_") %>%
  ggplot(aes(x = treatment, y = log2cpm, fill = replicate)) + geom_boxplot() +
  geom_hline(yintercept = median(mydgelist_cpm), lty = 2)

res.pca <- prcomp(t(mydgelist_cpm))
colors <- as.factor(str_split(rownames(mydgelist$samples), "_", simplify = TRUE)[,1])
fviz_pca_ind(res.pca,
             col.ind = colors,
             palette = "Set1",
             repel = TRUE,    # Avoid text overlapping
             title = "PCA_plot_of_WT_vs_LgDel_CNgV")
#ggsave("PCA_plot_of_WT_vs_LgDel_CNgV.png")

fviz_eig(res.pca, addlabels = TRUE)
fviz_contrib(res.pca, choice = "var", axes = 1, top = 100)


plotMDS(mydgelist,col= c("red","blue")[colors])
legend("topright",fill=c("red","blue"),legend=levels(colors))
#ggsave("MDS_PCA_plot_of_WT_vs_LgDel_CNgV.png")

most_var_genes <- sort(apply(mydgelist_cpm, 1, var), decreasing=TRUE)[1:500]
mydgelist_cpm_most_var <- mydgelist_cpm[most_var_genes,]

#pheatmap(mydgelist_cpm_most_var, show_rownames  = FALSE) # hiding rownames as there are 500 unreadable ones

my_z_score <- function(x){
  return( (x - mean(x)) / sd(x))
}

mydgelist_cpm_most_var2 <-  t(apply(mydgelist_cpm_most_var, 1, my_z_score))
#pheatmap(mydgelist_cpm_most_var2, show_rownames  = FALSE)
#ggsave("clustering_by_500_most_var_genes.png")

mydgelist_cpm2 <-  t(apply(mydgelist_cpm, 1, my_z_score))
#pheatmap(mydgelist_cpm2, show_rownames  = FALSE)
#ggsave("clustering_by_all_genes.png")
```

```{r incorporate batch effect into mydegelist_(edgeR manual) compute correlation between groups and design matrix}

mydgelist$samples$norm.factors
cpm1 <- cpm(mydgelist, log=FALSE)

mydgelist <- calcNormFactors(mydgelist)
mydgelist$samples$norm.factors
cpm2 <- cpm(mydgelist, log=FALSE)
group <- colors
#group = relevel(group, ref="WT")
rep <- factor(c(1,2,3,4,5,6,1,2,3,4,5,6))
Time <- factor(c("1", "1", "1", "2", "3", "4", "4", "5", "5", "6","6","6")) 
## This code is to just check the correlations between different batches
#design <- model.matrix(~ 0 + Time + Time : group)
#logFC <- predFC(mydgelist,design,prior.count=1,dispersion=0.05)
#cor(logFC[,5:8])

#This is the design for the Differential expression analysis
#design <- model.matrix(~group)
design <- model.matrix(~0+ Time + group)
#colnames(design) <- levels(group)
design

```



Robust glmlrt - less stringent
```{r estimate dispersion and perform robust generalized linear modeling with mydeglist_all genes}
mydgelist$genes <- data.frame(ENSEMBL=rownames(mydgelist))

mydgelist <- estimateGLMRobustDisp(mydgelist, design)
mydgelist$common.dispersion
plotBCV(mydgelist)

fit <- glmFit(mydgelist, design=design)
#plotQLDisp(fit)


lrt <- glmLRT(fit)
out1 <- topTags(lrt, n=Inf, sort.by = "PValue")
out1 <- as.data.frame(out1[["table"]])


out1 <- out1 %>%
  rename("ensembl_gene_id" = "ENSEMBL")

summary(out1)
```



```{r estimate dispersion and perform robust generalized linear modeling with mydeglist_all genes}
#*************************************************************************#

mydgelistTPM <- DGEList(counttablematrix)
mydgelistTPM <- calcNormFactors(mydgelistTPM)
mydgelistTPM$samples$norm.factors
mydgelistTPM$genes <- data.frame(ENSEMBL=rownames(mydgelistTPM))
mydgelistTPM <- estimateGLMRobustDisp(mydgelistTPM, design)


normalized_counts <- cpm(mydgelistTPM)

Gene_lengths <- read_csv("~/Downloads/Gene_lengths.csv")
Gene_lengths <- merge(normalized_counts, Gene_lengths, by.x=0, by.y="ensembl_gene_id")
Gene_lengths <- Gene_lengths %>%
  dplyr::select(c(1,12))

Gene_lengths <- Gene_lengths %>%
  pivot_wider(names_from = Row.names, values_from = gene_length)
(length(Gene_lengths) == nrow(normalized_counts))

Gene_lengths <- as.vector(Gene_lengths) %>% as.numeric()
tpm <- tpm.direct(normalized_counts, geneLength = Gene_lengths)
tpm <- data.frame(ensembl_gene_id = row.names(tpm), tpm)

tpm <- tpm %>%
  mutate(LgDel_NAC_mean = rowMeans(dplyr::select(.,c(2:6)))) %>%
  mutate(WT_mean = rowMeans(dplyr::select(.,c(7:11))))


```




```{r estimate dispersion and perform basic generalized linear modeling with mydeglist 10000 genes}

mydgelist$genes <- data.frame(ENSEMBL=rownames(mydgelist))

mydgelist <- estimateDisp(mydgelist, design, robust=TRUE)
mydgelist$common.dispersion
plotBCV(mydgelist)

fit <- glmFit(mydgelist, design)
#plotQLDisp(fit)


lrt <- glmLRT(fit)
out2 <- topTags(lrt, n=Inf, sort.by = "PValue")
out2 <- as.data.frame(out2[["table"]])

out2 <- out2 %>%
  rename("ensembl_gene_id" = "ENSEMBL")

summary(out2)

```




##glmQLFit - more stringent
```{r estimate dispersion and perform quasi-likelihood negative binomial generalized log-linear model with mydeglist genes}

mydgelist$genes <- data.frame(ENSEMBL=rownames(mydgelist))

mydgelist <- estimateDisp(mydgelist, design, robust=TRUE)
mydgelist$common.dispersion
plotBCV(mydgelist)

fit <- glmQLFit(mydgelist, design, robust=TRUE)
plotQLDisp(fit)


qlf <- glmQLFTest(fit)
out3 <- topTags(qlf, n=Inf, sort.by = "PValue")

out3 <- as.data.frame(out3[["table"]])
out3 <- out3 %>%
  rename("ensembl_gene_id" = "ENSEMBL")

summary(out3)

```

## Annotations using biomart with final gene table
```{r Annotate out table using mydgelist column names and biomart write xlsx table}


  ensembl <-  useDataset("mmusculus_gene_ensembl",mart=useMart("ensembl"))
  

#searchAttributes(mart = ensembl, pattern ="location")


  annotations <-  getBM(attributes = c("ensembl_gene_id", "mgi_id","mgi_symbol", "description", "gene_biotype", "chromosome_name"), 
                        filters = "ensembl_gene_id", 
                        values=rownames(mydgelistTPM), 
                        mart=ensembl, 
                        useCache = FALSE)
  
 #genelength <- getGeneLengthAndGCContent(annotations$ensembl_gene_id, "mm10", mode = "org.db")

anno_TPM <- merge(tpm, annotations, by = c("ensembl_gene_id" = "ensembl_gene_id"), all.x = TRUE)
PNs_5WTvs5LgDel_NAC <- merge(out1, anno_TPM, by = c("ensembl_gene_id" = "ensembl_gene_id"), all.y = TRUE)

#now merge tpm files, write table, remove duplicate ID cloumns manually then read again and merge with annotated glmlrt output from above
# tpmfiles <- dir(path = "/Users/shahrukh/Downloads/cortical_culture_rnaseq/TPM_filesLD_NACvsWT", 
#              pattern = "*.csv", full.names = T)
# 
# 
# tpmtable <- tpmfiles %>%
#   map(read_csv, col_names = TRUE, show_col_types = TRUE) %>%
#   reduce(inner_join, by = "ensembl_gene_id") %>%
#   group_by(ensembl_gene_id) %>%
#   filter(n_distinct(across(everything())) == 1) %>%
#   ungroup()

# EdgeR_glmlrt_TPM_4WTvs3LD_NAC <- inner_join(anno_glmlrt, tpmtable, by="ensembl_gene_id", all.x = TRUE)

write_xlsx(PNs_5WTvs5LgDel_NAC,'PNs_5WTvs5LgDel_NAC.xlsx')



```

